---
title: "Assignment 4"
author: "John Viviani"
date: "2/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How Did the AFC West Standings Change Over the Course of the 2019 NFL Regular Season?

```{r}

c("dplyr",
  "tidyverse",
  "ggthemes",
  "ggimage",
  "RCurl",
  "gridExtra",
  "plotly") -> package_names
for(package_name in package_names) {
  if(!is.element(package_name, installed.packages()[,1])) {
    install.packages(package_name,
                     repos = "http://cran.mtu.edu/")
  }
  library(package_name, character.only = TRUE,
          quietly = TRUE,verbose=FALSE)
}
rm(list=c("package_name", "package_names")) # clean up environment
options(scipen = 999)

```

The dataset used can be found here: https://github.com/fivethirtyeight/data/tree/master/nfl-elo

The dataset I am using consists of game-by-game data for NFL games data back to 1920. I filtered for all of the games during the 2019 NFL regular season, and also filtered for only the teams that are in the AFC West division (Denver Broncos, Kansas City Chiefs, Los Angeles Chargers, and the Oakland Raiders).

Since this dataset doesn't reveal which team won each game; I decided to manually add this to the dataset by reviewing the schedules of each of the 4 teams on nfl.com.

After creating a 'Win' column, as well as a column denoting which week during the regular season the game was played, I needed to create a 'Rank' column to show where each team was placed in the standings after each respective week. I manually added this based on the records of each team throughout the season. After this, I was able to create a visualization using plotly that showed where each team stood throughout the 17 week schedule of the 2019 NFL regular season.

```{r}
# Reading in the data
read.csv("C:/Users/John.JVivs/Documents/COLLEGE/GRAD SCHOOL/GDAT515/nfl_elo.csv") -> nfl

# Filtering for the 2019 NFL Regular Season
nfl[16275:16530,] -> nfl2019

str(nfl2019)

```


```{r}
# Filtering for the variables of interest
# Data is organized by game; therefore, collecting all home team stats in one data frame and all away team stats into another data frame to later combine
nfl2019 %>%
  select(., "date", "team1", "elo1_pre", "elo1_post") -> nfl2019_team1

nfl2019 %>%
  select(., "date", "team2", "elo2_pre", "elo2_post") -> nfl2019_team2

# Giving both data frames a common column name to bind later
nfl2019_team1$team <- nfl2019_team1$team1
nfl2019_team2$team <- nfl2019_team2$team2

# Binding the two data frames
newdf_2019 <- bind_rows(nfl2019_team1, nfl2019_team2)

```


```{r}
newdf_2019 %>%
  select(., 1,3,4,5,7,8) -> newdf_2019

# Same thing for pre and post qb values
newdf_2019$elo1_pre[is.na(newdf_2019$elo1_pre)] <- newdf_2019$elo2_pre[is.na(newdf_2019$elo1_pre)]

newdf_2019$elo1_post[is.na(newdf_2019$elo1_post)] <- newdf_2019$elo2_post[is.na(newdf_2019$elo1_post)]

# Arranging by date (Start from week 1 to week 4)
newdf_2019 %>%
  arrange(., date) %>% 
  select(., 1:4) %>%
  rename(., team_code = team, elo_pre = elo1_pre, elo_post = elo1_post) -> newerdf_2019

# Filtering for AFC West
newerdf_2019 %>%
  filter(., team_code == "DEN" | team_code == "KC" | team_code == "LAC" | team_code == "OAK") %>%
  arrange(., team_code) -> afcwest_2019

# Adding wins: 0 = Loss, 1 = Win
afcwest_2019 %>%
  mutate(., Win = c(0,0,0,0,1,1,0,0,1,0,0,1,1,0,1,1,1,1,1,1,0,0,1,0,1,0,1,1,1,1,1,1,1,0,0,1,0,0,0,1,1,0,0,0,1,0,0,0,1,0,0,1,1,0,0,1,1,1,0,0,0,0,1,0)) -> afcwest_2019

# Adding a cumulative sum of wins
afcwest_2019 %>%
  group_by(team_code) %>%
  mutate(cumsum = cumsum(Win)) -> afcwest_2019

# Creating a column that has respective week of the 2019 NFL regular season
c("1","2","3","4","5","6","7","8","9","11","12","13","14","15","16","17","1","2","3","4","5","6","7","8","9","10","11","13","14","15","16","17","1","2","3","4","5","6","7","8","9","10","11","13","14","15","16","17","1","2","3","4","5","7","8","9","10","11","12","13","14","15","16","17") -> Week

as.data.frame(Week) -> Week
as.character(Week$Week) -> Week$Week

# Attaching Week column to main dataframe
cbind(afcwest_2019, Week = Week$Week) -> afcwest_2019

# Creating a Rank column to show where a team stands in the division
c(4,4,4,4,4,3,3,4,4,4,4,3,3,3,3,2,2,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,3,3,3,3,3,4,4,3,3,3,3,4,4,4,4,4,1,2,2,2,2,2,2,2,2,2,2,2,2,2,2,3) -> Rank

as.data.frame(Rank) -> Rank
as.character(Rank$Rank) -> Rank$Rank

cbind(afcwest_2019, Rank = Rank$Rank) -> afcwest_2019
```

## Data Visualization with Plotly

```{r}
plot_ly(data = afcwest_2019, x = Week, y = ~cumsum, color = ~team_code, colors = c("Orange", "Red", "Cyan", "Black"), mode = "lines+markers", text = ~paste0("Wins: ", cumsum, "<br>Week: ", Week, "<br>Rank: ", Rank)) %>%
  layout(xaxis = list(ticks = "Outside",
                      title = "Week",
                      showgrid = FALSE,
                      zeroline = FALSE,
                      showticklabels=FALSE),
         yaxis = list(ticks = "Outside",
                      dtick = 1,
                      ticklen = 16,
                      tickwidth = 1,
                      tickcolor = toRGB("red"),
                      title = "Wins",
                      showgrid = FALSE,
                      zeroline = FALSE),
         showlegend = TRUE,
         legend = list(x = .2, y = .85),
         title = "NFL 2019 Regular Season: AFC West Wins by Team")

```

The data visualization shows where each team ranked in the division over the course of the season. By hovering over each point, you can see where each team was ranked during the respective week of the season, as well as how many wins the team had at that point in time. The teams are toggleable in the legend as a means of seeing data points that overlap one another (e.g. Raiders and Chargers overlap in Week 4). By hovering over the last point for each team, you are able to see each team's final rank in the AFC West standings for the 2019 NFL regular season.
